White and Red Wine Quality by Jennifer Joo
========================================================

For this project, we will look at a dataset of red and white verde wine samples.  The variables in the dataset include:

fixed acidity 
volatile acidity 
citric acid 
residual sugar 
chlorides 
free sulfur dioxide 
total sulfur dioxide 
density 
pH 
sulphates 
alcohol
quality (score between 0 and 10)
type (red, white)

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

# install.packages("ggplot2", dependencies = T) 
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)
 
library(ggplot2)
library(knitr)
library(dplyr)
library(corrplot)
library(GGally)

```

```{r global_options, include=FALSE}
# prevent warnings from showing in knitted file
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, Load_the_Data}
# data source: http://archive.ics.uci.edu/ml/datasets/Wine+Quality
# my wd: C:/Users/jjoo/Desktop/udacity_data_analyst_projects/p4-explore-data-in-r

# Load the Data
red <- read.csv("winequality-red.csv", sep = ";")
red$type <- 'red'
white <- read.csv("winequality-white.csv", sep = ";")
white$type <- 'white'

data <- rbind(red, white)

```

# Univariate Plots Section

We will begin the analysis by looking at the distribution of each of the above variables individually to see if they're skewed, contain outliers, etc.  We'll also take a quick look at the overall structure of the data and some summary statistics (mean, median, quartiles)

```{r echo=FALSE, Univariate_Plots}
# look at the data frame
str(data)

# look at basic summary statistics for this dataset
summary(data)
```

We have a categorical variable and 12 numeric/integer variables. The single categorical variable is 'type', which indicates whether a datapoint is associated is a red or white wine.

```{r echo=FALSE}
# hist plots for fixed.acidity, volatile.acidity, citric.acid, and residual.sugar
fixed.acidity_hist <- ggplot(data, aes(x=fixed.acidity)) +
          geom_histogram(colour="#2eaaff", fill="#6666FF") +
          ggtitle("Fixed Acidity Histogram")

fixed.acidity_hist

```
The fixed acidity histogram is approximately normal with a slight right skew with a mean of about 6.
```{r echo=FALSE}
volatile.acidity_hist <- ggplot(data, aes(x=volatile.acidity)) +
          geom_histogram(colour="#333333", fill="#f4b90c", bins = 70) +
          ggtitle("Volatile Acidity Histogram")

volatile.acidity_hist
```

The fixed acidity histogram is approximately normal with a right skew with a mean of about 0.25.

```{r echo=FALSE}
citric.acid_hist <- ggplot(data, aes(x=citric.acid)) +
          geom_histogram(colour="#333333", fill="#ed7aff") +
          ggtitle("Sulphates Histogram")
citric.acid_hist 
```
The citric acid histogram is approximately normal with a right skew with a mean of about 0.25.


```{r echo=FALSE}
residual.sugar_hist <- ggplot(data, aes(x=residual.sugar)) +
          geom_histogram(colour="#333333", fill="#01ba55") +
          ggtitle("Residual Sugar Histogram")

residual.sugar_hist
```
The citric acid histogram is somewhat normal with a right skew with a mean of about 2.


```{r echo=FALSE}
# hist plots for chlorides, free.sulfur.dioxide, total.sulfur.dioxide, and density
chlorides_hist <- ggplot(data, aes(x=chlorides)) +
          geom_histogram(colour="#333333", fill="#b4006d") +
          ggtitle("Chlorides Histogram") +
chlorides_hist
```

The chlorides histogram is approximately normal with a long right tail.

```{r echo=FALSE}
free.sulfur.dioxide_hist <- ggplot(data, aes(x=free.sulfur.dioxide)) +
          geom_histogram(colour="#333333", fill="#728f00") +
          ggtitle("Free Sulfur Dioxide Histogram")
free.sulfur.dioxide_hist
```
The free sulfur dioxide histogram is approximately normal with an outlier at ~290.

```{r echo=FALSE}
total.sulfur.dioxide_hist <- ggplot(data, aes(x=total.sulfur.dioxide)) +
          geom_histogram(colour="#333333", fill="#ff4887") +
          ggtitle("Total Sulfur Dioxide Histogram")
total.sulfur.dioxide_hist

```
The total sulfur dioxide histogram has two peaks at around 20 and 120 with a few outliers above 300.

```{r echo=FALSE}
density_hist <- ggplot(data, aes(x=density)) +
          geom_histogram(colour="#333333", fill="#00b0a3") +
          ggtitle("Density Histogram")
density_hist
```
The density histogram is somewhat normal with outliers at around 1.01 and 1.04

```{r echo=FALSE}
pH_hist <- ggplot(data, aes(x=pH)) +
          geom_histogram(colour="#333333", fill="#a31927") +
          ggtitle("pH Histogram")
pH_hist
```
The pH histogram is approximately normal; it's interesting to note that all wines have a pH of about 4 or lower so all wines are at least somewhat acidic.

```{r echo=FALSE}
sulphates_hist <- ggplot(data, aes(x=sulphates)) +
          geom_histogram(colour="#333333", fill="#476697") +
          ggtitle("Sulphates Histogram")
sulphates_hist
```
The sulphates histogram is approximately normal with a right skew.  Most values are less than 1.0

```{r echo=FALSE}
alcohol_hist <- ggplot(data, aes(x=alcohol)) +
          geom_histogram(colour="#333333", fill="#ffb35c") +
          ggtitle("Alcohol Histogram")
alcohol_hist
```
The alcohol histogram is less normal looking than our other graphs; the largest peak is around 7 with a couple of smaller peaks at ~10 and ~14.
```{r echo=FALSE}
quality_hist <- ggplot(data, aes(x=quality)) +
          geom_bar(colour="#333333", fill="#b464c0") +
          ggtitle("Quality Histogram")
quality_hist
```

The quality histogram is approximately normal; most wines have a rating of 6 and the second most common rating is a 5.  The unevent distribution here may affect our analysis later on since there are few wines rated a 3 or 9.

```{r echo=FALSE}
type_hist <- ggplot(data, aes(x=type)) +
    geom_bar(colour="#333333", fill="#6b7fce") +
    ggtitle("Wine Type Histogram")
type_hist
```

There are about 3 times as many white wines as red wines, which is important to note since it will affect how later multivariate graphs may be perceived.

# Univariate Analysis

### What is the structure of your dataset?

There are 6,497 observations and 13 variables.  All but one variable ('type') have a data type of numeric or integer.  Most of the numeric/integer variables have an approximately normal distribution and most variables have some outliers.

### What is/are the main feature(s) of interest in your dataset?

The main features of interest are quality, which is a score between 0 and 10, and type, a created variable that indicates whether the wine is red or white.  However, it is important to note that available quality scores range from 3 to 9.  The quality instagram is approximately normal although ideally, we would like an equal distribution of scores across both red and white wines.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Without looking further at the data set, I think that pH and alcohol may be important.

### Did you create any new variables from existing variables in the dataset?

Yes, I merge the red wine and white wine datasets and created the variable "type" to indicate which dataset the observations came from.  The possible values are "red" and "white".  It's important to note that this dataset is unbalanced; there are about 3 times as many white wines as red wines in this dataset.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

There are more white wine observations than red wine observations and this may affect the data analysis and any plots to follow but I didn't adjust this distribution, since I didn't want to exclude any data from the analysis.  The distribution for alcohol had many peaks, which I thought was unusual but if I had to hypothesize, I would say that maybe wines were more likely to have certain alcohol levels or this distribution is due to the type of wine (white or red).  The distribution for sulphates also had a sudden peak at around 5.0; this may be due to how the data was collected.


# Bivariate Plots Section

We'll first create a box plot by quality and wine type to see if quality ratings for red and white wines vary drastically.

```{r echo=FALSE, Bivariate_Plots}
# box plot -- quality by type

box_plot1 <- ggplot(data, aes(x=type, y=quality, fill=type)) +
  geom_boxplot() + ggtitle("Box Plot for Quality by Wine Type") +
  labs(subtitle="White Wine has one more outlier than red red
                but otherwise, they have about the same distribution")

# box plot -- pH by type
box_plot2 <- ggplot(data, aes(x=type, y=pH, fill=type)) +
          geom_boxplot() + ggtitle("Box Plot for pH by Wine Type")+
          labs(subtitle="Red wine generally has a larger pH than white wine")

box_plot3 <- ggplot(data, aes(x=type, y=alcohol, fill=type)) +
          geom_boxplot() + ggtitle("Box Plot for Alcohol by Wine Type")+
          labs(subtitle="White wine generally has a slightly higher alcohol value than red")

print (box_plot1); print (box_plot2); print (box_plot3)

print('pH Quartiles for red and white wine: ')
print(quantile(red$pH))
print(quantile(white$pH))

print('Alcohol Quartiles for red and white wine: ')
print(quantile(red$alcohol))
print(quantile(white$alcohol))

```

```{r echo=FALSE, Bivariate_Plots}
# correlation matrix
# for reference, see:
# https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
df_corr <- subset( data, select=-c(type) )
mcor <- cor(df_corr)
round(mcor, digits=2)

cex.before <- par("cex")
par(cex = 0.7)
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(mcor, method="color",
            shade.col=NA, tl.col="black", tl.srt=45, order="FPC",
            col=col(50), addCoef.col="black", addcolorlabel="yes")
par(cex = cex.before)


```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

In the box plots, quality didn't vary much by wine type; the majority of the datapoints had a quality rating of 5 or 6, although there white wines had an outlier that had a quality of 9.  This may be due to the fact that very few wines were rated a 3 or 9.

However, it is interesting to note that red wines were generally less acidic (had a higher pH value) than white wines.  White wines also had more outliers in this box plot distribution although there are also more observations for white wines, so this I wouldn't consider this too unusual.  White wines had a median pH of 3.18 and red wines had a median pH of 3.31, with a difference of 0.13.

White wines also had a somewhat higher alchol content and a larger interquartile range for alcohol level as well.  White wines had a median alcohol level of 10.2 and red a median alcohol level of 10.4.


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?

The strongest relationship was between total.sulfur.dioxide and free.sulfur.dioxide which has a correlation of 0.72, which made sense since they were both measures of sulfur dioxide.  The next strongest was between density and residual.sugar with a correlation of 0.55.

The two variables that had the largest negative correlation was alcohol and density.  The most interesting relationship was between quality and alcohol, which had a correlation coefficient of 0.44.


# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

# ggpairs -- part 1
# I split the columns up because the graph was getting too big
# I chose the variables that I thought were most interesting
ggpairs(data,
        columns=c('density', 'pH', 'sulphates', 'alcohol', 'quality'), # columns to include in plot
        title='Wine GGpairs plot - selected columns', # title of the plot
        mapping=ggplot2::aes(colour = type)) # how to segment the plot
```
For pH, sulphates, alcohol, and quality, red and white wines have similar distribution shapes although for pH and sulphates, red wines have a somewhat larger pH and sulphares median, which we saw in the bivariate plots section.

Red and white wines have different distributions when we look at density.  Red wines have a density distribution that is tall and narrow while white wines have a density distribution that is shorter and wider.  We also see two consistent outliers for white wines when we compare density to pH, sulphates, alcohol, and quality. 

```{r echo=FALSE, Multivariate_Plots}
# bar graphs
ggplot(data = data, aes(y = pH, x = factor(type))) +
    geom_jitter(alpha = 0.4) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    ggtitle("Bar Plot for alcohol segmented by quality and wine type")
```

Wine wines are overall more acidic and more dense than white wines but there are also more white wine data points than red wine data points so this is expected.

```{r echo=FALSE, Multivariate_Plots}
# scatter plot by wine type
ggplot(data,aes(x=pH, y=alcohol, colour=type)) +
    geom_point(alpha=.3) +
    geom_smooth(alpha=.25, method = "loess", span = 0.9) +
    ggtitle("pH and Alcohol Scatterplot by Wine Type")

```

There isn't a clear negative or positive correlation between alcohol and pH either for white wine.  The line for red wine slants upward after a pH of 3.5 but this may be due to outliers.

```{r echo=FALSE, Multivariate_Plots}
ggplot(data, aes(x = pH, y = alcohol, colour=type)) +
  geom_point() +
  geom_density2d()

```

The centers for red wines and white wines for the pH vs alcohol graph are very close to one another.  The alcohol value for each center is around 7 and the pH center for white wines is around 3.2 and for red wines, it is around 3.6.  We saw in an earlier box plot that white wines are slightly more acidic than white wines.

```{r echo=FALSE, Multivariate_Plots}
# scatter grid
data_selectQual <- subset(data, quality >= 5 & quality <= 7)

ggplot(data_selectQual, aes(x=pH, y=alcohol, colour = factor(quality))) +
  geom_point() + stat_density2d() +
  facet_grid(quality ~ type) +
  labs(title = 'pH vs alcohol scatterplot (factored by type and quality)',
      subtitle = 'Subsetted by quality ratinfs of 5,6,7')

```

There appears to be weak correlation between pH and alcohol for both red and white wines at quality ratings of 5, 6, and 7.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

The relationship between alcohol and pH was positive for red wines at a pH level greater than 3.5.  Overall, the relationship between pH and quality did not have a very positive relationship for white wines.  However, for red wines, there was a slight negative relationship between quality and pH for red wines.

When we subset the pH vs alcohol scatterplot by quality ratings of 5,6, and 7 and wine type, it's interesting to note that centers are in approximately the same location for each scatterplot. I would have expected the centers would change based on quality, alcohol, and pH level since I do somewhat expect that a wine rated a 5 would be different somehow to a wine rated a 7.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(data, aes(x=type, y=pH, fill=type)) +
          geom_boxplot() + ggtitle("Box Plot for pH by Wine Type")+
          labs(subtitle="Red wine generally has a larger pH than white wine") +
    xlab('wine type') + ylab('pH level')
  
```

### Box Plot for pH by Wine Type
Red wines were somewhat less acidic (had a higher pH value) than white wines.  White wines also had more outliers in this box plot distribution although there are also more observations for white wines, so this is expected.  White wines had a median pH of 3.18 and red wines had a median pH of 3.31, with a difference of 0.13.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(data,aes(x=pH, y=alcohol, colour=type)) +
    geom_point(alpha=.3) +
    geom_smooth(alpha=.25, method = "loess", span = 0.9) +
    ggtitle("pH and Alcohol Scatterplot by Wine Type") +
    labs(subtitle = 'For white wines, pH vs. alcohol stays relatively steady,
                but for red wines, alcohol starts to increast at a pH of 3.5') +
    ylab('alcohol (% ABV)') + xlab('pH level')


```

### pH and Alcohol Scatterplot by Wine Type
There isn't a clear negative or positive correlation between alcohol and pH either for white wine.  The line for red wine slants upward after a pH of 3.5 but this may be due to outliers.

### Plot Three
```{r echo=FALSE, Plot_Three}

data_selectQual <- subset(data, quality >= 5 & quality <= 7)

ggplot(data_selectQual, aes(x=pH, y=alcohol, colour = factor(quality))) +
  geom_point() + stat_density2d() +
  facet_grid(quality ~ type) +
  labs(title = 'pH vs alcohol scatterplot (factored by type and quality)',
      subtitle = 'Subsetted by quality ratings of 5,6,7') +
      ylab('alcohol (% ABV)') + xlab('pH level')

```

### pH and Alcohol scatterplot (factored by type and quality of 5, 6, and 7)

When we subset the pH vs alcohol scatterplot by quality ratings of 5,6, and 7 and wine type, we see that there is one center for all scatterplots except for white wines that were rated a 7, although these centers are relatively close to one another; it's also interesting to note that centers are in approximately the same location for each scatterplot.

------

# Reflection
One of the difficult things about this dataset was that there were quite a few columns and I didn't really know how to start looking at it, especially since my knowledge of wine is non-existent. Also, since most wines were rated a 5, 6, or 7, it was somewhat difficult to see what qualities would rate a wine a 9; as we can see in plot three, there are no red wines that were rated a 9 so I couldn't do a comparison there.  Otherwise, I enjoyed analysing this dataset and I think the large number of numerical data made this easier to analyse than say, a dataset with purely categorical values.

For future work, I would also like to attempt to predict a wine's quality rating based on the other provided variables.  Not one variable had a very strong correlation with quality. Alcohol had the highest correlation value with quality at 0.44.