White and Red Wine Quality by Jennifer Joo
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

# install.packages("ggplot2", dependencies = T) 
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)
 
library(ggplot2)
library(knitr)
library(dplyr)
library(corrplot)

```

```{r echo=FALSE, Load_the_Data}
# data source: http://archive.ics.uci.edu/ml/datasets/Wine+Quality

# Load the Data
wd = "C:/Users/joonn/Desktop/udacity_data_analyst_projects/p4-explore-data-in-r/"
setwd(wd)

red <- read.csv("winequality-red.csv", sep = ";")
red$type <- 'red'
white <- read.csv("winequality-white.csv", sep = ";")
white$type <- 'white'

data <- rbind(red, white)

str(data)
summary(data)

```

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}

par(mfrow=c(3,2))

ggplot(data, aes(x=quality)) + geom_bar()
ggplot(data, aes(x=chlorides)) + geom_bar()
ggplot(data, aes(x=sulphates)) + geom_bar()
ggplot(data, aes(x=pH)) + geom_bar()
ggplot(data, aes(x=volatile.acidity)) + geom_bar()

# box plot
ggplot(data, aes(x=type, y=quality)) + geom_boxplot()
ggplot(data, aes(x=type, y=pH)) + geom_boxplot()

```

# Univariate Analysis

### What is the structure of your dataset?

There are 6,497 observations and 13 variables.

### What is/are the main feature(s) of interest in your dataset?

The main features of interest are quality, which is a score between 0 and 10, and type, a created variable that indicates whether the wine is red or white.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Without looking further at the data set, I think that pH, quality, and alcohol may be important.

### Did you create any new variables from existing variables in the dataset?

Yes, I merge the red wine and white wine datasets and created the variable "type" to indicate which dataset the observations came from.  The possible values are "red" and "white".

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

The distribution for chlorides showed a right skew.


# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

# density plots by wine type (red, white)
ggplot(data, aes(x=quality, colour=type)) + geom_density()

# box plot -- quality by type
ggplot(data, aes(x=type, y=quality, fill=type)) + geom_boxplot()


# correlation matrix
# for reference, see:
# https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
df_corr <- subset( data,select=-c(type) )
mcor <- cor(df_corr)

library(corrplot)
col <- colorRampPalette( brewer.pal( 6 , "RdBu" ) )

corrplot(mcor, method="shade",
            shade.col=NA, tl.col="black", tl.srt=35, type="lower",
            col=col(200), addCoef.col="black", addcolorlabel="no")


```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

In the density plot, it appears as though white wines are rated more highly but since there are more white wines, this graph may be misleading.  Looking at the box plot, the majority of each wine (red and white) are rated between a 5 and 6 but white wines have more outliers that are rated more highly.


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?

The strongest relationship was between total.sulfur.dioxide and free.sulfur.dioxide which has a correlation of 0.72.  The next strongest was between density and residual.sugar with a correlation of 0.55.




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

# bar graphs
ggplot(data, aes(x=quality, y=alcohol, fill=type)) + geom_bar(stat="identity")
ggplot(data, aes(x=pH, y=alcohol, fill=type)) + geom_bar(stat="identity")

# scatter and lin regression lines
ggplot(data, aes(x=pH, y=quality, color=type)) + geom_point(shape=1) +
    scale_colour_hue(l=75) +
    geom_smooth(method=lm,   # Add linear regression lines
                se=FALSE, fullrange=TRUE) # extend the regression lines

multiplot(p1, p2, p3, p4, cols=2)

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
